Biology background

### 2014-10-29 Rmarkdown/biology_background.Rmd
End of day commit

Sir Adrian Bird, an esteemed British geneticist and someone who is in favour of the broader definition, attempts to unite these definitions in the following ([@Bird:2007em]):

@@

## DNA methylation

DNA methylation is a chemical modification of DNA that can impart information on top of the DNA sequence. It is heritable during mitotic cell division, which means that it is faithfully copied across to the daughter cell during cell division[^dnam_copying]. It therefore fits into Art Riggs' definition of an epigenetic modification [@russo1996epigenetic]. So pervasive is DNA methylation that [@Lister:2009boa] dubbed it the "fifth base" of the DNA code.

@@

What is referred to as DNA methylation is typically methylation of the cytosine nucleotide. Cytosine methylation is by far the most common form of DNA methylation in the animal and plant kingdoms[^other_DNAm], however, I will continue to use the terms _DNA methylation_ and _cytosine methylation_ interchangeably as is standard in the literature.

[^other_DNAm]: Two additional forms of DNA methylation, N6-Methyladenine (m6A) and N4-methylcytosine (m4C), are primarily found in prokaryotes (__SOURCE: http://en.wikipedia.org/wiki/DNA_methyltransferase#cite_note-21__) (or is it bacterial (__SOURCE: http://www.ncbi.nlm.nih.gov/pubmed/16479578__))

A German chemist, W.G. Ruppel, first identified a methylated nucleic acid in 1898. Ruppel was studying _tuberculinic acid_, the poison of _Mycobacterium tuberculosis_[^tuberculosis] , and discovered that it contained a methylated nucleotide (__SOURCE__). In 1925, Johnson and Coghill isolated this methylated nucleotide as a product of hydrolysis of tuberculinic acid [@Johnson:1925js]. However, Johnson and Coghill's results were disputed for over twenty years by other researchers who were unable to replicate the original findings [@VISCHER:1949ty].

In 1945, Hotchkiss ultimately proved Johnson and Coghill correct when he isolated 5-methylcytosine from nucleic acid prepared from cow thymus [@HOTCHKISS:1948va]. Using paper chromatography, Hodgkiss demonstrated that methylated cytosine existed and was distinct from conventional cytosine and uracil.

@@

In the next section I focus on DNA methylation in mammals, in particular humans, but include brief descriptions of DNA methylation in other organisms.

### DNA methylation in mammals
DNA methylation is essential for normal development in mammals, such as human and mouse (__REFERENCE__). My thesis focuses on DNA methylation in humans (_Homo sapiens_) but much of the material in this section is true for other mammals.

The typical site of DNA methylation is at the C5 carbon position of a cytosine nucleotide. This modified cytosine is called _5-Methylcytosine_ or _5mC_. [__FIGURE__](http://en.wikipedia.org/wiki/File:5-Methylcytosine.svg) shows the structure of 5mC.

At the level of single-stranded DNA, methylation is a binary event: a cytosine is either methylated or unmethylated. For double-stranded DNA, (__TODO: Clarify; only CG methylation can be strand symmetric. CHG and CHH are not strand symmetric motifs__) methylation is usually symmetric across the two strands (_strand symmetric_) although _hemimethylation_ does occur. Within a diploid cell a particular cytosine may be unmethylated or methylated on both homologous chromosomes or methylated on one chromosome and unmethylated on the other. While the former that is more common, examples of the latter case, such as _allele-specific methylation_ and _genomic imprinting_, are of considerable biological interest (__SOURCE/EXAMPLES__).

In mammalian genomes, most cytosines are unmethylated except for those at the _CpG dinucleotide_, which is usually strand-symmetric. I explain the importance of CpG dinucleotides in the next section.

### CpG dinucleotides

A _CpG dinucleotide_, or more simply a _CpG_, is a cytosine followed by a guanine in the linear DNA sequence (__FIGURE__). The "p" stands for the phosphate backbone of DNA and some authors omit it in favour of simply calling it CG methylation. Approximately 60-80% of CpGs are methylated in mammals, meaning that the cytosine in the dinculeotide is a 5-Methylcytosine (__REFERENCE__). I will use the abbreviation _mCpG_ to refer to a methylated CpG, _uCpG_ to refer to an unmethylated CpG and simply _CpG_ when the methylation state is irrelevant.

@@

One reason for this relative scarcity of CpGs is that methylated cytosines can spontaneously deaminate to thymines ([@Scarano:1967ue]). Thus, over time, many mCpGs will become TpGs, leading to a genome-wide reduction in the proportion of CpGs and a genome-wide increase in the proportion of TpGs (see __FIGURE__).

There are many other evolutionary pressures on the distribution of nucleotides in a genome. One effect of this is that the distribution of CpGs is far from uniform. I discuss this distribution in the next section.

### CpG islands and other sandy metaphors

@@

The 20-40% of CpGs that are methylated in mammalian genomes are mostly found in clusters called _CpG islands_ (_CGIs_). Again, there are evolutionary arguments for why these CGIs form (__REFERENCE__).

CGIs are important regulatory elements in the genome and are where most differences in DNA methylation between different cell types are found (__REFERENCE__). [@GardinerGarden:1987we] gave the classical definition of a CGI:

> For the purpose of this survey, regions of DNA with a moving average of %G+C over 50 and Obs/Exp CpG over 0.6 have been classed as CpG-rich regions. CpG-rich regions over 200 bp in length are unlikely to have occurred by chance alone, so, as a working definition, have been labelled as CpG islands.

Other authors have since provided alternative definitions (__REFERENCES__ Adrian Bird, Rob Klose, Hao Wu, etc.)

[@Wu:2010do] developed a Hidden Markov Model to define CGIs. In my thesis I have used the predicted CGIs from this model. These predicted CGIs are available from [http://rafalab.jhsph.edu/CGI/](http://rafalab.jhsph.edu/CGI/).

The "sandy/beachy" metaphors have been continued (i.e. stretched to breaking point) with various authors defining CGI shores, CGI shelves, CpG canyons, CpG deserts and CpGs in the "open sea". CGI shores [@Irizarry:2008hg] are defined as regions within 2kb of CGIs. They regions have been demonstrated to have an increased variability of CpG methylation [@Irizarry:2008hg]. CGI shelves (__WHO__) are defined as __WHAT__ and __WHY ARE THEY USEFUL__. CpG deserts are __WHAT__, CpG canyons are __WHAT__ and the "open sea" is __WHAT AND HOW DOES THIS DIFFER FROM CpG DESERTS__.

### Non-CpG methylation

@@

### Modifications of a modification

@@

### Writers, readers and erasers

@@

### Function of DNA methylation: "normal", cancer and development

### DNA methylation in other organisms and kingdoms

#### Insects {-}

@@

#### Plants {-}

@@

#### Fungi {-}

#### Bacteria {-}

#### Archaea {-}

## Assays for studying DNA methylation

@@

### Enzyme digestion assays
_Restriction endonucleases_ are an important technique in molecular biology. These enzymes can preferrentially "cut/cleave/digest" DNA at particular sequence motifs. The motif at which a restriction enzyme cleaves DNA is called the _recognition motif_ or _restriction sequence_. The methylation of a position in the recognication motif can inhibit a restriction enzyme from cleaving the DNA. This can be used to design an assay to infer the methylation state of a DNA fragment.

@@

### Affinity enrichment assays
Affinity enrichment assays compare measurements between an "enriched" version and an "input" (control) version of the same sample to infer the presence or absence of DNA methylation. This is very similar to the idea behind chromatin immunoprecipitation followed by microarray hybridisation (ChIP-chip) or sequencing (ChIP-seq). Enrichment may be done using a methylation-specific antibody, such as __EXAMPLE__, or a __unmethylation-specific(?)__ antibody, such as __EXAMPLE__.

@@

### Sodium bisulfite conversion assays
In the 1980s, __Hayatsu OR Wang et al.__ discovered that when _denatured_ (__DEFINE?__) DNA is treated with sodium bisulfite (__CHEMICAL SYMBOL__), unmethylated cytosines deaminate to uracils much faster than do methylated cytosines (__SOURCE__). The methylated cytosines are said to be "protected" from conversion to uracils. When the bisulfite treated is amplified by PCR the uracils are subsequently converted to thymines.

@@

#### Analysis platforms combined with bisulfite conversion {-}

@@

### DNA kinetics assays
__TODO__

### Pros and cons of bisulfite-conversion assays
Bisulfite-conversion assays are considered the "gold standard" for studying DNA methylation. Currently, multiple cells are required as input although the minimum number required is being lowered and efforts are well produce single-cell, whole-genome DNA methylation maps using bisulfite-treated DNA (__SOURCE__).

@@

## Whole-genome bisulfite-sequencing

@@

### Targeted BS-seq, e.g. RRBS, NimbleGen and Agilent's capture kits, MRE-seq

>>>>

Sir Adrian Bird, an esteemed British geneticist and someone who is in favour of the broader definition, attempts to unite these definitions in the following \cite{Bird:2007em}:

@@

\section{DNA methylation}\label{sec:dna_methylation}

DNA methylation is a chemical modification of DNA that can impart information on top of the DNA sequence. It is heritable during mitotic cell division, which means that it is faithfully copied across to the daughter cell during cell division[^dnam_copying]. It therefore fits into Art Riggs' afore-mentioned definition of an epigenetic modification \cite[p1]{russo1996epigenetic}. DNA methylation is found in bacteria, fungi, plants and animals.

@@

A major reason that DNA methylation is studied is that it is essential for normal development in mammals \cite{Li:1992bp}. It is also involved in many key biological processes related to human development and health, such as the regulation of gene expression (__CITE__), silencing of transposable elements (__CITE__), X chromosome inactivation (__CITE__) and tumorigenesis (__CITE__). However, DNA methylation is not found in all organisms, for example, it not detectable in yeast, and so is not essential for life. Furthermore, the level of DNA methylation varies widely amongst different organisms, with many having very low levels of methylation (__TODO: examples and citations__).

When people speak of DNA methylation they are generally referring to methylation of the cytosine nucleobase. Even more specifically, they are referring to the molecule _5-methylcytosine_. 5-methylcytosine, abbreviated as _5mC_, is by far the most common form of DNA methylation in the animal and plant kingdoms[^other_DNAm]. However, I will continue to use the terms _DNA methylation_ and _cytosine methylation_ interchangeably, as is standard in the literature. The importance of 5-methylcytosine in mammalian genomes is such it has been dubbed the "fifth base" of the DNA code \cite{@Lister:2009boa}.

[^other_DNAm]: Two additional forms of DNA methylation are N6-methyladenine (m6A) and N4-methylcytosine (m4C). N6-methyladenine is a methylated form of adenine, which is found in mRNA \cite[e.g.,]{Adams:1975ki} and DNA, although the latter only in bacterial DNA \cite{Ratel:2006dv}. N4-methylcytosine has also been detected in bacterial DNA \cite[e.g.,]{Ehrlich:1985bv, Ratel:2006dv}.

A German chemist, W.G. Ruppel, first identified a methylated nucleic acid in 1898. Ruppel was studying _tuberculinic acid_, the poison of _Mycobacterium tuberculosis_[^tuberculosis] , and discovered that it contained a methylated nucleotide (__TODO: Can only find first page of this paper and it's in German. Need to verify__). In 1925, Johnson and Coghill isolated 5-methylcytosine as a product of hydrolysis of tuberculinic acid, the nucleic acid of _Mycobacterium tuberculosis_ \cite{Johnson:1925js}. However, Johnson and Coghill's results were disputed for over twenty years by other researchers who were unable to replicate the original findings \cite{@VISCHER:1949ty}.

@@

In 1945, Hotchkiss ultimately proved Johnson and Coghill correct when he isolated _5-methylcytosine_ from nucleic acid prepared from cow thymus \cite{HOTCHKISS:1948va}. Using paper chromatography, Hotchkiss demonstrated that methylated cytosine existed and was distinct from conventional cytosine and uracil.

\subsection{DNA methylation in mammals}\label{sec:dna_methylation_in_mammals}

- [x] Done

DNA methylation is essential for normal development in mammals, such as human and mouse (__REFERENCE__). My thesis focuses on DNA methylation in humans but much of the material in this section is true for other mammals.

The typical site of DNA methylation is at the C5 carbon position of a cytosine nucleotide. This modified cytosine is called _5-Methylcytosine_ or _5mC_. [__FIGURE__](http://en.wikipedia.org/wiki/File:5-Methylcytosine.svg) shows the structure of 5mC. It is common to describe a cytosine as being "methylated" or "unmethylated", however, care must be taken in defining these terms.

At the lowest level, the level of single-stranded DNA, methylation is a binary event: a cytosine is either methylated or unmethylated. Double-stranded DNA, at least at _palindromic_ methylation loci[^palindromic_sequence], is generally symmetrically methylated, i.e., the loci on each strand are both methylated or both unmethylated. However, _hemimethylation_, where the methylation loci on one strand is methylated and its partner on the opposite strand is unmethylated, can and does occur.

[^palindromic_sequence]: "A palindromic sequence is a nucleic acid sequence (DNA or RNA) that is the same whether read 5' (five-prime) to 3' (three prime) on one strand or 5' to 3' on the complementary strand with which it forms a double helix" (__TODO: current source is wikipedia, find and properly cite__).

Within a diploid cell, a particular cytosine may be unmethylated or methylated on both homologous chromosomes or methylated on one chromosome and unmethylated on the other. While the former that is more common, examples of the latter case, such as _allele-specific methylation_ and _genomic imprinting_, are of considerable biological interest (__TODO: sources and examples__).

In mammalian genomes, most cytosines are unmethylated except for those at _CpG dinucleotides_. A _CpG dinucleotide_, or more simply a _CpG_, is a cytosine followed by a guanine in the linear DNA sequence (__FIGURE__). The "p" stands for the phosphate backbone of DNA and some authors omit it in favour of simply calling it CG methylation. A CpG is a palindromic sequence and is generally symmetrically methylated. Approximately 60-80% of CpGs are methylated in mammals, meaning that the cytosine in the dinculeotide is a 5-methylcytosine (__REFERENCE__).

\subsection{CpG dinucleotides}\label{cpg_dinucleotides}

- [x] Done

@@

One reason for this relative scarcity of CpGs is that methylated cytosines can spontaneously deaminate to thymines \cite{@Scarano:1967ue}. Thus, over time, many methylated CpGs will become TpGs, leading to a genome-wide reduction in the proportion of CpGs and a genome-wide increase in the proportion of TpGs (see __FIGURE__).

There are many other evolutionary pressures on the distribution of nucleotides in a genome. One effect of this is that the distribution of CpGs is far from uniform. In fact, CpGs tend to form clusters, which are termed CpG islands.

\subsection{CpG islands and other sandy metaphors}\label{sec:cpg_islands}

- [ ] Done

@@

The 20-40% of CpGs that are methylated in mammalian genomes are mostly found in clusters called _CpG islands_ (_CGIs_). Again, there are evolutionary arguments for why these CGIs form (__TODO: sources__).

CGIs are important regulatory elements in the genome and are where most differences in DNA methylation between different cell types are found (__TODO: sources__). The classical definition of a CpG island, given by \cite{GardinerGarden:1987we} and used by, for example, the UCSC genome browser, is a region of the genome where:

1. The moving average of GC-content is greater than $50\%$, and
2. The observed-to-expected ratio of CpGs is greater than $0.6$, and
3. The region is longer than 200 bp.

This definition was refined by \cite{Takai:2002jb} to exclude _Alu_-repetitive elements, which otherwise are misclassified as _bona fide_ CpG islands. More recently, \cite{@Wu:2010do} developed a hidden Markov model to predict CpG islands based on the CpG density and GC-content of the region.

An alternative definition, and one that pre-dates the definition of \cite{GardinerGarden:1987we}, is based on the identification of unmethylated regions of the genome, which are typically CpG-dense. Such regions were previously called HpaII tiny fragment islands, or _HTF islands_, and named after the restriction enzyme used to identify them \cite{Cooper:1983ev, Bird:1985gj}.

The "sandy/beachy" metaphors have been continued (i.e., stretched to breaking point) with various authors defining CpG island shores, CpG island shelves, CpG canyons and CpGs in the "open sea":

- CpG island shores: regions within 2 kb of CpG islands \cite{Irizarry:2008hg}. CpG island shores have been demonstrated to have an increased variability of CpG methylation \cite{Irizarry:2008hg}.
- CpG island shelves have been defined as regions within 2 kb of a CpG island shore \cite{Bibikova:2011fw}.

__UP TO HERE__

- CpG canyons are __WHAT__ and the
- "open sea" is __WHAT AND HOW DOES THIS DIFFER FROM CpG DESERTS__.
- methylation deserts: 100 Kbp windows with the lowest 1% methylation level in the sample of interest

Another metaphor, based on methylation levels rather than CpG density, are methylation deserts. These regions are sample-dependent and were defined by \cite{Li:2012gg} as 100 Kbp windows with the lowest 1% methylation level in the given sample.

\subsection{Non-CpG methylation}\label{sec:non_cpg_methylation}

- [ ] Done

@@

\subsection{Modifications of a modification}\label{sec:modifications_of_a_modification}

- [ ] Done

@@

\subsection{Writers, readers and erasers}\label{sec:writers_reads_and_erasers}

@@

\subsection{Function of DNA methylation: "normal", cancer and development}\label{sec:function_of_dna_methylation}

- [ ] Done

\subsection{DNA methylation in other organisms}\label{sec:dna_methylation_in_other_organisms}

- [ ] Done

__TODO: Don't want too much detail here__

\subsubsection{Insect}

- [ ] Done

@@

\subsubsection{Plants}

- [ ] Done

@@

\subsubsection{Fungi}

- [ ] Done

\subsubsection{Bacteria}

- [ ] Done

\subsubsection{Archaea}

- [ ] Done

\section{Assays for studying DNA methylation}\label{sec:dna_methylation_assays}

@@

\subsection{Enzyme digestion assays}\label{sec:enzyme_digestion_assays}

- [ ] Done

_Restriction endonucleases_ are an important technique in molecular biology. These enzymes can preferrentially "cut/cleave/digest" DNA at particular sequence motifs. The motif at which a restriction enzyme cleaves DNA is called the _recognition motif_ or _restriction sequence_. The methylation of a position in the recognication motif can inhibit a restriction enzyme from cleaving the DNA. This can be used to design an assay to infer the methylation state of a DNA fragment.

@@

\subsection{Affinity enrichment assays}\label{sec:affinity_enrichment_assays}

- [ ] Done

Affinity enrichment assays compare measurements between an "enriched" version and an "input" (control) version of the same sample to infer the presence or absence of DNA methylation. This is very similar to the idea behind chromatin immunoprecipitation followed by microarray hybridisation (ChIP-chip) or sequencing (ChIP-seq). Enrichment may be done using a methylation-specific antibody, such as __EXAMPLE__, or a __unmethylation-specific(?)__ antibody, such as __EXAMPLE__.

@@

\subsection{Sodium bisulfite conversion assays}\label{sec:bs_assays}

- [ ] Done

In the 1980s, __Hayatsu OR Wang et al.__ discovered that when _denatured_ (__DEFINE?__) DNA is treated with sodium bisulfite (__CHEMICAL SYMBOL__), unmethylated cytosines deaminate to uracils much faster than do methylated cytosines (__SOURCE__). The methylated cytosines are said to be "protected" from conversion to uracils. When the bisulfite treated is amplified by PCR the uracils are subsequently converted to thymines.

@@

\subsubsection{Analysis platforms combined with bisulfite conversion}\label{sec:bs_platforms}

- [ ] Done

@@

\subsection{DNA kinetics assays}\label{sec:dna_kinetics_assays}

- [ ] Done

\subsection{Pros and cons of bisulfite-conversion assays}\label{sec:pros_and_cons_of_bs_assays
}

- [ ] Done

Bisulfite-conversion assays are considered the "gold standard" for studying DNA methylation. Currently, multiple cells are required as input although the minimum number required is being lowered and efforts are well produce single-cell, whole-genome DNA methylation maps using bisulfite-treated DNA (__SOURCE__).

@@

\section{Whole-genome bisulfite-sequencing}\label{sec:wgbs}

@@

\subsection{Targeted BS-seq, e.g. RRBS, NimbleGen and Agilent's capture kits, MRE-seq}\label{sec:targeted_bsseq}

- [ ] Done
